\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-05-01 latex/ms.tex
add better real injections

In \autoref{fig:issue2}, we simulate a light curve featuring a single transit in addition with stellar variability of different timescales and amplitudes. In case \textit{a} (purple in \autoref{fig:issue2}), we model and detrend these signals using the widely-adopted Tukey's biweight filtering method presented in \citealt{tukey} and its implementatation from \texttt{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}. In case \textit{b} (red in \autoref{fig:issue2}) we employ the same GP model used to simulate stellar variability to reconstruct and remove it from light curves ignoring the presence of potential transits (the case of unknown transits search). Once transits cleaned from stellar variability, we asses their remaining SNR using \autoref{eq:snr}.
\\\\
\autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and suggests that this degradation is strongly dependant on the stellar variability characteristics encountered. In order to explore the parameter space for which detrending is the most problematic, we simulate light curves including a single transit and a much wider range of correlated noise characteristics defined with two parameters: $\tau$ the relative timescale of the variability with respect to the transit duration; and $\delta$, the relative amplitude of the variability against transit depth, such that:
\begin{equation}\label{eq:relative_param_space}
    \tau \propto \frac{\mathrm{variability\; timescale}}{\mathrm{transit\;duration}} \quad \text{and} \quad 
    \delta \propto \frac{\mathrm{variability\; amplitude}}{\mathrm{transit\;depth}} 
\end{equation}
To follow this parametrization, we simulate the photometric variability using a GP with an SHO kernel with hyperparameters \footnote{}:
\begin{equation}\label{eq:relative_params}
    \omega = \frac{\pi}{\tau D} \quad 
    \sigma = \delta \frac{\Delta}{2} \quad  \text{and}  \quad  
    Q = 10
\end{equation}
with $\Delta$ and $D$ the depth and duration of the simulated transit, similar in all light curves. 
% We fix a relatively high value for the quality factor $Q$ in order to restrict our simulations to strongly periodic variability signals. 
For $\tau=1$ and $\delta=1$, the expressions of $\omega$ and $\sigma$ given in \autoref{eq:relative_params} correspond to a periodic signal with a period half that of the transit duration, and a peak to peak amplitude two times that of the transit depth, i.e. strongly resembling the simulated transit signal. As in \autoref{fig:issue2}, we reconstruct the variability using a biweight filter with a window length of $3D$ (three times the transit duration, an optimal value according to \citealt{wotan}). We then subtract the variability from each light-curve, estimate the resulting transit depth (using the in-transit minimum flux) and compute its SNR. \autoref{fig:snr_detrend} shows the remaining SNR values of transits computed this way, after each variability signals with random $(\tau, \delta)$ have been detrended.
\\\\
\autoref{fig:snr_detrend} shows that it exists an entire region of the $(\tau, \delta)$ parameter space for which detrending degrades transit SNR to the point of no detection ($SNR < 6$). While this region might only represent a fraction of existing exoplanetary systems, we argue that these systems are of great value for the field of Exoplanetary Science. Indeed, as variability is often linked to the presence of starspots, systems whose  with the stellar photo

@@

\newpage
\section{Validation with injection-recovery}\label{simu}
In order to validate our algorithm, we assess the transit search capabilities of \texttt{nuance} against commonly used techniques. As demonstrated in \autoref{detrending_effect}, the full-fledged modeling offered by \texttt{nuance} is not always necessary, and becomes beneficial for particular noise characteristics that are relative to the searched transit parameters. Hence, our comparison will be done on the relative parameter space $(\tau, \delta)$ described in \autoref{eq:relative_param_space}, focusing on its formulation by assuming correlated noise in the form of stellar variability (\autoref{eq:relative_params}) simulated using a Gaussian Process with an SHO kernel (\autoref{app_gp}). Ultimately, we demonstrate in this section how \texttt{nuance} compares with commonly used techniques, depending on the similarity between stellar variability and the searched transit signals.
\\\\
In what follows, we inject transits in simulated or real datasets, and employ algorithms to recover these transits with the right parameters (hence \textit{injection-recovery}). We consider the transit to be found if the absolute difference between the periods $dP$ and epochs $dT_0$ of the recovered and injected planets follow
$$dP<0.1\;\text{days} \hspace{0.5cm} \text{and} \hspace{0.5cm} dT_0<0.01\;\text{days}$$

@@

\subsection{Comparison with \wtls{} on a simulated light curves}

We start by comparing \texttt{naunce} to a relatively simple method on simulated light curves. As in \autoref{detrending_effect}, this simple method consist in searching for planetary transits by first modeling stellar variability using an optimal biweight filter (implemented in \texttt{wõtan} \citep{wotan} using a window size three times that of the injected transit duration), subtracting it from the light-curve, and then perform the search on the detrended light curve using an optimal box-least-square algorithm (\texttt{TLS}, \citealt{tls}). We name this method \wtls{}.
\\\\
The dataset consists in 10 000 light curves simulated using the model described in \autoref{app_principle_simulations}. We simulate a common periodic transit among all light curves, of period $P=1.1$ days, epoch $T_0=0.2$ days, duration $D=0.04$ days and depth $\Delta=2\%$. Each light curve consist in a 3 days observation, with an exposure time of 2 minutes, leading to $N=2160$ data points with a normal error of $0.1\%$.
\\\\
For a given pair of $(\tau, \delta)$, we simulate stellar variability using a Gaussian Process with an SHO kernel of hyperparameters defined by \autoref{eq:relative_params}, computed with respect to the injected transit parameters $D$ and $\Delta$. The same kernel is used for the search with \texttt{nuance}, an optimal choice on equal footing with the optimal biweight filter employed in the \wtls{} search. We generate 10 000 pairs of $(\tau, \delta)$ such that
$$\tau \sim \mathcal{U}(0.1, 40)\hspace{0.5cm}\text{and}\hspace{0.5cm} \delta \sim \mathcal{U}(0.1, 100)$$
where $\mathcal{U}(a, b)$ denotes a uniform distribution of lower bound $a$ and upper bound $b$.
\\\\
The results of these injection-recovery are shown in \autoref{fig:simu} and highlight particularly well the benefit of \nuance{} against \wtls{} on transits corresponding to $\delta > 3$ and $\tau > 2$, i.e. with relatively low transit depth compared to the stellar variability amplitude, and a relatively small duration compared to the stellar variability period. On average, searching a single light curve with \nuance{} took 9 seconds, compared to 8 seconds with \wtls{}. Beyond its increased sensitivity, this metric demonstrates the tractability of \nuance{}, studied in more details in \autoref{}.
\\\\
About these results, we note these transit searches were done in a particularly optimal setup, on light curves not all physically realistic, hence demonstrating the performance of \nuance{} on a theoretical basis only. In the next subsection, we perform transits injection-recovery on a real dataset, using more realistic tools and on light curves featuring known planetary transits, hence validating \nuance{} on a realistic multi-planetary dataset.

@@

>>>>

In \autoref{fig:issue2}, we simulate a light curve featuring a single transit in addition with stellar variability of different timescales and amplitudes. In case \textit{a} (purple in \autoref{fig:issue2}), we model and detrend these signals using the widely-adopted Tukey's biweight filtering method presented in \citealt{tukey} and its implementatation from \texttt{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}. In case \textit{b} (red in \autoref{fig:issue2}) we employ the same GP model used to simulate stellar variability to reconstruct and remove it from light curves ignoring the presence of potential transits (the case of unknown transits search). Once transits cleaned from stellar variability, we asses their remaining SNR using \autoref{eq:snr}.
\\\\
\autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and suggests that this degradation is strongly dependant on the stellar variability characteristics encountered. In order to explore the parameter space for which detrending is the most problematic, we simulate light curves including a single transit and a much wider range of correlated noise characteristics defined with two parameters: $\tau$ the relative timescale of the variability with respect to the transit duration; and $\delta$, the relative amplitude of the variability against transit depth, such that:
\begin{equation}\label{eq:relative_param_space}
    \tau \propto \frac{\mathrm{variability\; timescale}}{\mathrm{transit\;duration}} \hspace{0.5cm} \text{and} \hspace{0.5cm} 
    \delta \propto \frac{\mathrm{variability\; amplitude}}{\mathrm{transit\;depth}}.
\end{equation}
To follow this parametrization, we simulate the photometric variability using a GP with an SHO kernel with hyperparameters \footnote{}:
\begin{equation}\label{eq:relative_params}
    \omega = \frac{\pi}{\tau D}, \hspace{0.5cm} 
    \sigma = \delta \frac{\Delta}{2} \hspace{0.5cm}  \text{and}  \hspace{0.5cm}  
    Q = 10,
\end{equation}
with $\Delta$ and $D$ the depth and duration of the simulated transit, similar in all light curves. 
% We fix a relatively high value for the quality factor $Q$ in order to restrict our simulations to strongly periodic variability signals. 
For $\tau=1$ and $\delta=1$, the expressions of $\omega$ and $\sigma$ given in \autoref{eq:relative_params} correspond to a periodic signal with a period half that of the transit duration, and a peak to peak amplitude two times that of the transit depth, i.e. strongly resembling the simulated transit signal. As in \autoref{fig:issue2}, we reconstruct the variability using a biweight filter with a window length of $3D$ (three times the transit duration, an optimal value according to \citealt{wotan}). We then subtract the variability from each light-curve, estimate the resulting transit depth (using the in-transit minimum flux) and compute its SNR. \autoref{fig:snr_detrend} shows the remaining SNR values of transits computed this way, after each variability signals with random $(\tau, \delta)$ have been detrended.
\\\\
\autoref{fig:snr_detrend} shows that it exists an entire region of the $(\tau, \delta)$ parameter space for which detrending degrades transit SNR to the point of no detection ($SNR < 6$). While this region might only represent a fraction of existing exoplanetary systems, we argue that these systems are of great value for the field of Exoplanetary Science. Indeed, as variability is often linked to the presence of starspots, systems whose  with the stellar photo

@@

\newpage
\section{Validation on simulated light curves}\label{simu}
In order to validate our algorithm, we assess the transit search capabilities of \texttt{nuance} against commonly used techniques. As demonstrated in \autoref{detrending_effect}, the full-fledged modeling offered by \texttt{nuance} is not always necessary, and becomes beneficial for particular noise characteristics that are relative to the searched transit parameters. Hence, our comparison will be done on the relative parameter space $(\tau, \delta)$ described in \autoref{eq:relative_param_space}, focusing on its formulation by assuming correlated noise in the form of stellar variability (\autoref{eq:relative_params}) simulated using a Gaussian Process with an SHO kernel (\autoref{app_gp}). Ultimately, we demonstrate in this section how \texttt{nuance} compares with commonly used techniques, depending on the similarity between stellar variability and the searched transit signals.
\\\\
In what follows, we inject transits in simulated or real datasets, and employ algorithms to recover these transits with the right parameters (hence \textit{injection-recovery}). We consider the transit to be found if the absolute difference between the periods $dP$ and epochs $dT_0$ of the recovered and injected planets follow
$$dP<0.1\;\text{days} \hspace{0.5cm} \text{and} \hspace{0.5cm} dT_0<0.01\;\text{days}$$

@@

We start by comparing \texttt{naunce} to a relatively simple method on simulated light curves. As in \autoref{detrending_effect}, this simple method consist in searching for planetary transits by first modeling stellar variability using an optimal biweight filter (implemented in \texttt{wõtan} \citep{wotan} using a window size three times that of the injected transit duration), subtracting it from the light-curve, and then perform the search on the detrended light curve using an optimal box-least-square algorithm (\texttt{TLS}, \citealt{tls}). We name this method \wtls{}.
\\\\
The dataset consists in 10 000 light curves simulated using the model described in \autoref{app_principle_simulations}. We simulate a common periodic transit among all light curves, of period $P=1.1$ days, epoch $T_0=0.2$ days, duration $D=0.04$ days and depth $\Delta=2\%$. Each light curve consist in a 3 days observation, with an exposure time of 2 minutes, leading to $N=2160$ data points with a normal error of $0.1\%$.
\\\\
For a given pair of $(\tau, \delta)$, we simulate stellar variability using a Gaussian Process with an SHO kernel of hyperparameters defined by \autoref{eq:relative_params}, computed with respect to the injected transit parameters $D$ and $\Delta$. The same kernel is used for the search with \texttt{nuance}, an optimal choice on equal footing with the optimal biweight filter employed in the \wtls{} search. We generate 10 000 pairs of $(\tau, \delta)$ such that
$$\tau \sim \mathcal{U}(0.1, 40)\hspace{0.5cm}\text{and}\hspace{0.5cm} \delta \sim \mathcal{U}(0.1, 100)$$
where $\mathcal{U}(a, b)$ denotes a uniform distribution of lower bound $a$ and upper bound $b$.
\\\\
The results of these injection-recovery are shown in \autoref{fig:simu} and highlight particularly well the benefit of \nuance{} against \wtls{} on transits corresponding to $\delta > 3$ and $\tau > 2$, i.e. with relatively low transit depth compared to the stellar variability amplitude, and a relatively small duration compared to the stellar variability period. On average, searching a single light curve with \nuance{} took 9 seconds, compared to 8 seconds with \wtls{}. Beyond its increased sensitivity, this metric demonstrates the tractability of \nuance{}, studied in more details in \autoref{}.
\\\\
About these results, we note these transit searches were done in a particularly optimal setup, on light curves not all physically realistic, hence demonstrating the performance of \nuance{} on a theoretical basis only. In the next section, we perform transits injection-recovery on a real dataset, using more realistic tools and on light curves featuring known planetary transits, hence validating \nuance{} on a realistic multi-planetary dataset.

@@

\section{Validation on real TESS light curves}\label{real}

The parameter space ($\tau$, $\delta$) presented in is convenient from a signal analysis standpoint, but does not easily translate into physical cases. In order to assess the performance of \texttt{nuance} on real light curves, we rather evaluate its transit recovery capabilities in the planetary (period, radius) parameter space, more often encountered in the literature (e.g. ).
